---
title: "Getting started with the `calibrar` package"
author: "Ricardo Oliveros-Ramos"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with the `calibrar` package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE}
library(calibrar)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`calibrar` provides general-purpose optimisation tools in R, with an `optim()`-compatible
interface and additional functionality that is particularly useful when objective functions are
computationally expensive. Although the package was originally developed for parameter
estimation (calibration) of complex ecological and simulation models, its core optimisation
functions can be used to minimise any user-defined objective function.

The package exposes three complementary entry points:

- `optim2()`: an extension of `stats::optim()` with a curated set of optimisers and additional
  features such as parameter masking (`active`), optional parallel finite-difference gradients,
  and support for structured (list-based) parameters.
- `optimh()`: a unified interface to a broad collection of heuristic optimisation methods, with
  standardised argument naming.
- `calibrate()`: a higher-level workflow for sequential (multi-phase) calibration, where
  parameter masking can be progressively relaxed across phases, optionally with replicates for
  stochastic models.

In addition, `calibrar` includes helper tools to build objective functions from simulated
outputs and observational data. These are optional: advanced users can supply objective
functions directly to `optim2()`, `optimh()` or `calibrate()`.

## Basic usage

This vignette introduces three front ends. Choose the one that matches your task:

- General optimisation with an `optim()`-like interface (plus masking / parallel finite differences /
  list parameters): use `optim2()`.
- Heuristic optimisation across many algorithms with standardised argument naming: use `optimh()`.
- Sequential (multi-phase) calibration for complex or stochastic models, with progressive
  parameter unmasking: use `calibrate()`.

### optim2(): general-purpose optimisation (an extension of `stats::optim()`)

`optim2()` uses the same calling convention as `stats::optim()` for the common case of a flat numeric parameter vector:

```{r, echo=TRUE, results='markup'}
library(calibrar)

f <- function(x) sum(x^2)

o1 <- stats::optim(par = rep(1, 5), fn = f)
o2 <- optim2(par = rep(1, 5), fn = f)

o1$value; o2$value
o1$par;   o2$par
```

The results are identical, as here `optim2()` acts just as a wrapper for `stats::optim()`. In addition, `optim2()` can work with structured (list-based) parameters, which can simplify objective functions for complex models:

```{r, echo=TRUE, results='markup'}
par0 <- list(curve = list(a = 1, b = 0.5), offset = 0)

fn <- function(par) {
(par$curve$a - 2)^2 + (par$curve$b - 1)^2 + (par$offset - 0.1)^2
}

out <- optim2(par = par0, fn = fn)
out$par
```

The main arguments of `optim2()` are:

```{r, echo=TRUE, eval=FALSE, results='markup'}
optim2(
  par,
  fn,
  gr = NULL,
  ...,
  method = c("Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN", "Brent", "nlm", "nlminb",
    "Rcgmin", "Rvmmin", "hjn", "spg", "LBFGSB3", "AHR-ES"),
  lower = -Inf,
  upper = +Inf,
  active = NULL,
  control = list(),
  hessian = FALSE,
  parallel = FALSE
)
```

The first difference is the possible values for the `method` argument. In addition to the first six methods, also available in `optim()`, `optim2()` gives access to `stats::nlm()` and `stats::nlminb()` but with the same syntax as `optim()` to make them easy to use. In addition, three methods from the `optimr` package (`Rcgmin`, `Rvmmin`, `hjn`), the L-BFGS-B v3 implemented in the `bfgsb3c` package and the `AHR-ES` (Adaptative Hierarchical Recombination Evolutionary Strategy) implemented in this package.

We can run the same previous example with two other methods not available to `optim()`:

```{r, echo=TRUE, results='markup'}
optim2(par=rep(1, 5), fn=function(x) sum(x^2), method="nlm")
```

```{r, echo=TRUE, results='markup'}
set.seed(880820) # for reproducibility
optim2(par=rep(1, 5), fn=function(x) sum(x^2), method="AHR-ES")
```

The second difference is the new argument `active`, which is a vector indicating if a parameter will be optimized (i.e. active) or fixed to a constant value during the optimization process. In the next example, we will fix the third and fourth parameters to its initial values:

```{r, echo=TRUE, results='markup'}
optim2(par=rep(1, 5), fn=function(x) sum(x^2), 
       active=c(TRUE, TRUE, FALSE, FALSE, TRUE))
```

As we can see, in the final solution, the `par` value keep the values at 1, the initial value provided. All the numerical gradients computed internally have also 'masked' this parameters and the derivatives are not computed for them to speed up computation time.

Finally, the third difference is the new argument `parallel`, that activates the parallel computation of the numerical gradient, when `gr` is not supplied:

```{r, echo=TRUE, results='markup'}
optim2(par=rep(1, 5), fn=function(x) sum(x^2), parallel=TRUE)
```

This last option will increase performance when the computation time of `fn` in considerable. We will explain in detail this feature in a following section.

Additionally, the method for the computation of the numerical gradient can be chosen within the `control` list:

```{r, echo=TRUE, results='hide', eval=FALSE}
optim2(par=rep(0.5, 5), fn=function(x) sum(2*x^(3.1*x)), control=list(gr.method="richardson"))
optim2(par=rep(0.5, 5), fn=function(x) sum(2*x^(3.1*x)), control=list(gr.method="central"))
optim2(par=rep(0.5, 5), fn=function(x) sum(2*x^(3.1*x)), control=list(gr.method="forward"))

```

### optimh()

The function `optimh()` has a similar functionality as `optim2()` but acts as a wrapper for several heuristic optimization algorithms implemented in several packages: `dfoptim`, `optimr`, `minqa`, `cmaes`, `genSA`, `DEoptim`, `soma`, `rgenoud` and `psoptim`. The `optimh()` function standardizes the inputs and outputs to those of `optim()`, providing a more convenient user interface. All specific arguments of this methods can be passed to the original function using the `control` argument.

```{r, echo=TRUE, eval=FALSE, results='markup'}
optimh(
  par,
  fn,
  gr = NULL,
  ...,
  method = c("AHR-ES", "Nelder-Mead", "SANN", "hjn", "CMA-ES", "genSA", "DE", "soma",
    "genoud", "PSO", "hybridPSO", "mads", "hjk", "hjkb", "nmk", "nmkb"),
  lower = -Inf,
  upper = +Inf,
  active = NULL,
  control = list(),
  hessian = FALSE,
  parallel = FALSE
)
```

```{r, echo=TRUE, results='markup'}
# Covariance Matrix Adaptation Evolutionary Strategy
set.seed(880820) # for reproducibility
optimh(par=rep(1, 5), fn=function(x) sum(x^2), method="CMA-ES",
       control=list(maxit=200))
```

```{r, echo=TRUE, results='markup'}
# Generalized Simulated Anneling
set.seed(880820) # for reproducibility
optimh(par=rep(1, 5), fn=function(x) sum(x^2), method="genSA", 
       lower=rep(-100, 5), upper=rep(100, 5),
       control=list(maxit=200, temperature=6000))
```

```{r, echo=TRUE, results='markup'}
# Self-Organising Migrating Algorithm
set.seed(880820) # for reproducibility
optimh(par=rep(1, 5), fn=function(x) sum(x^2), method="soma",
       lower=rep(-100, 5), upper=rep(100, 5),
       control=list(maxit=200))
```
The `maxit` control argument has been standardized to work with all methods and to represent the maximum number of iterations of the algorithm. However, the specific number of function evaluations per iteration may vary between methods. You can refer to the help pages from each package to have details about every specific method and its control arguments.

### Running in parallel

Most algorithms implemented in `optim2()` and some in `optimh` can benefit of parallel computation. For the methods that uses the numerical computation of the gradient, this will be calculated in parallel. In order to support any type of parallel implementation, the parallel setup is NOT automatic, and must be done by the user previous to executed the optimization, as described in the following example:

```{r, echo=FALSE, results='hide', eval=TRUE, message=FALSE}
library(parallel)
```

```{r, echo=TRUE, results='markup', eval=FALSE}
library(parallel)
ncores = detectCores() - 1 # number of cores to be used
cl = makeCluster(ncores)
# this is slower than sequential for very fast models (like this one)
optim2(par=rep(0.5, 5), fn=function(x) sum(x^2), 
               control=list(ncores=ncores), parallel=TRUE)
stopCluster(cl) # close the parallel connections
```

## Sequential parameter estimation

### calibrate()

The `calibrate()` function implements an automatic sequential parameter estimation, meaning parameters can be un-masked (set active) progressively during sequential phases of the calibration process.

```{r, echo=TRUE, eval=FALSE, results='markup'}
calibrate(
  par,
  fn,
  gr = NULL,
  ...,
  method = NULL,
  lower = NULL,
  upper = NULL,
  phases = NULL,
  control = list(),
  hessian = FALSE,
  replicates = 1,
  parallel = FALSE
)
```

With the basic syntax, the `calibrate()` function will work similarly to `optim2()` and `optimh()`, performing a simple optimization:

```{r, echo=TRUE, results='markup'}
calibrate(par=c(1,2,3,NA,NA), fn=function(x) sum(x^2))
```

If `upper` and `lower` bounds are provided, the calibrate function can take `NA` as starting values for the optimization, to some or all of the parameters:

```{r, echo=TRUE, results='markup'}
calibrate(par=c(1,2,3,NA,5), fn=function(x) sum(x^2),
          lower=rep(-100, 5), upper=rep(100, 5))
```

### Setting up a parameter estimation with multiple phases.

Multiple `phases` can be set up by selecting a different one for each parameter. The starting value of the optimization for each phase is updated with the best parameters found in the previous phase:

```{r, echo=TRUE, results='markup'}
calibrate(par=c(1,2,3,NA,5), fn=function(x) sum(x^2),
          lower=rep(-100, 5), upper=rep(100, 5),
          phases=c(1,2,3,2,1))
```

When a phase is set to a negative number, the parameter is fixed at its initial value during all the calibration and it is never optimized:

```{r, echo=TRUE, results='markup'}
calibrate(par=c(1,2,3,NA,5), fn=function(x) sum(x^2),
          lower=rep(-100, 5), upper=rep(100, 5),
          phases=c(1,2,-1,2,1))
```

### Dealing with stochastic functions

When dealing with stochastic functions, the argument `replicates` can be helpful, as it allows to evaluate the objective function several times, taking the average value of them as the actual function value (as an approximation of the expected value of the function). When `replicates=1`, the algorithm used by default is "LBFGSB3", but when replicates is greater than 1, "AHR-ES" is used. The next examples use the function `sphereN()`, which computes the Euclidean distance from a point `x` to the origin of coordinates after a random displacement of its position (see `?sphereN` for details). We will set the maximum number of iterations to 1000 to speed up the execution of the vignette.

```{r, echo=TRUE, results='markup'}
calibrate(par=c(1,2,3,NA,5), fn=sphereN,
          lower=rep(-100, 5), upper=rep(100, 5),
          phases=c(1,2,3,2,1), replicates=3, control=list(maxit=1000))
```

The number of replicates can be one single value or a vector with the length equal to the number of phases:

```{r, echo=TRUE, results='markup'}
calibrate(par=c(1,2,3,NA,5), fn=sphereN,
          lower=rep(-100, 5), upper=rep(100, 5),
          phases=c(1,2,3,2,1), replicates=c(1,1,5), control=list(maxit=1000))
```

### Parameters as lists

```{r, echo=TRUE, results='markup'}
calibrate(par=list(par1=c(1,2,3), par2=NA, par3=5), fn=sphereN,
          lower=rep(-100, 5), upper=rep(100, 5),
          phases=c(1,2,-3,2,1), replicates=c(1,5), control=list(maxit=1000))
```

Note that the function `fn` must be able to take a list as parameter set, and the user must ensure this works beforehand.

### Running in parallel

Most algorithms implemented in the `calibrate` function can benefit of parallel computation. For the methods that uses the numerical computation of the gradient, this will be calculated in parallel as in `optim2()`. In order to support any type of parallel implementation, the parallel setup is NOT automatic, and must be done by the user previous to executed the optimization, as described in the following example:

```{r, echo=TRUE, results='hide', eval=FALSE}
library(parallel)
ncores = detectCores() - 1 # number of cores to be used
cl = makeCluster(ncores)
# this is slower than sequential for very fast models (like this one)
calib = calibrate(par=rep(0.5, 5), fn=sphereN,
                  replicates=3, 
                  lower=rep(-5, 5), 
                  upper=rep(+5, 5), 
                  phases=c(1,1,1,2,3), 
                  control=list(parallel=TRUE, ncores=ncores))
stopCluster(cl) # close the parallel connections
```

While parallelising an optimization can speed up computations, it is not always faster than running the optimisation sequentially. Parallel execution introduces additional overhead, such as data communication, thread or process management, and synchronization, which can outweigh the benefits when the objective function is relatively fast to evaluate. As a result, for computationally inexpensive functions, the time required to coordinate parallel execution may lead to slower performance compared to a sequential approach. However, as the evaluation time of the objective function increases, the efficiency gains from parallelisation become more significant, making it a valuable strategy for complex or computationally demanding optimization problems.

Please, refer to `vignette(package="calibrar")` for additional vignettes or to the [calibrar website](https://roliveros-ramos.github.io/calibrar/) for more details.
