---
title: "Parameter estimation for ODE systems"
author: "Ricardo Oliveros-Ramos"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter estimation for ODE systems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette focus on the use of the `calibrate()` for parameter estimation of Ordinary Differential Equations (EDO) systems. We suggest to see the vignette 'Getting started with the `calibrar` package' before reading this one, specially if you do not have previous experience doing optimization in R. Reading the vignette 'Using the `calibrate()` function for parameter estimation' may be also very useful, as not every detail is reproduced here.

```{r, include=FALSE}
knitr::opts_current$get(c(
  "cache",
  "cache.path",
  "cache.rebuild",
  "dependson",
  "autodep"
))
```

## Fitting an EDO system: the Lotka-Volterra Predator-Prey model

We will estimate the parameters for a predator-prey Lotka-Volterra model using the `calibrate()` function. The model is defined by a system of ordinary differential equations for the abundance of prey $N$ and predator $P$:
$$\frac{dN}{dt} = rN(1-N/K)-\alpha NP$$
$$\frac{dP}{dt} = -lP + \gamma\alpha NP$$
The parameters to estimate are the prey’s growth rate $r$, the predator’s mortality rate $l$, the carrying capacity of the prey $K$ and $\alpha$ and $\gamma$ for the predation interaction. To start, we created the demonstration data for this model using the function `calibrar_demo()` function with `T=100` as an additional argument to specify the time horizon. 

```{r, cache=TRUE}
library(calibrar)
set.seed(880820)
path = NULL # NULL to use the current directory
LV = calibrar_demo(path=path, model='PredatorPrey', T=100) 
setup = calibration_setup(file = LV$setup)
observed = calibration_data(setup=setup, path=LV$path)
run_model = calibrar:::.PredatorPreyModel
coef(LV)
```
The `run_model` will simulate the data, by solving the ODE system defined by the Lotka-Volterra model:

```{r, eval=FALSE}
run_model = function(par, T) {
  if(!requireNamespace("deSolve", quietly = TRUE)) 
    stop("You need to install the 'deSolve' package.")
  # par is a list with 'alpha', 'beta' 'gamma', 'sd' and 'mu_ini'.
  LV = function(t, y, parms, ...) {
    r = parms$r
    l = parms$l
    alpha = parms$alpha
    gamma = parms$gamma
    K = parms$K
    dN = r*y[1]*(1-(y[1]/K)) - alpha*y[1]*y[2]
    dP = -l*y[2] + gamma*alpha*y[1]*y[2]
    return(list(c(dN, dP)))
  }
  times = seq(0, T)
  y0 = c(par$initial$N, par$initial$P)
  sol = deSolve::ode(y=y0, times=times, func=LV, parms=par, method="ode45")
  out = as.list(as.data.frame(sol[,-1]))
  names(out) = c("prey", "predator")
  out$prey[is.na(out$prey)] = 0
  out$predator[is.na(out$predator)] = 0
  return(out)
}
```

The core of the `run_model` function relies in solving the ODE system by using the `ode()` function of the `deSolve` package. In general, the `run_model` takes a `par` argument (that can be a vector or a list) and produce a named list with the simulated data. All the intermediate code solves the simulation problem of taking a set of parameters to produce numerical outputs for each simulated variable. We will also define the objective function, based on the setup created by the demo:

```{r, echo=FALSE, results='asis'}
knitr::kable(setup)
```
The `calibration_objFn()` will automatically create the objective function following the information in the setup, in this case, by using a log-normal distribution for the errors (`type`) and the same `weight` for both. The data is expected to be read from the files in the `file` column, files that were created when calling the demo. 

```{r}
# objective functions
obj = calibration_objFn(model=run_model, setup=setup, observed=observed, T=LV$T, aggregate=TRUE)
```

Now we can fit the model using several optimization methods:

```{r optim1, cache=TRUE}
lbfgsb1 = calibrate(par=LV$guess, fn=obj, , method='L-BFGS-B', lower=LV$lower, upper=LV$upper, phases=LV$phase)
lbfgsb2 = calibrate(par=LV$guess, fn=obj, lower=LV$lower, upper=LV$upper, phases=LV$phase, method="Rvmmin")
ahr = calibrate(par=LV$guess, fn=obj, method='AHR-ES', lower=LV$lower, upper=LV$upper, phases=LV$phase)
nm = calibrate(par=LV$guess, fn=obj, phases=LV$phase, method="Nelder-Mead")
```

And compare them:

```{r}
summary(LV, lbfgsb1, lbfgsb2, ahr, nm, show_par = 1:5)
```

When a function is created with the `calibration_objFn()` it gains a `predict()` method, that can be used to simulate the model with the estimated set of parameters.

```{r}
lbfgsb1.pred = predict(lbfgsb1)
lbfgsb2.pred = predict(lbfgsb2)
ahr.pred     = predict(ahr)
nm.pred      = predict(nm)
```

and plot the results.

```{r, fig.asp=0.66, fig.width=8}
methods = c("data", "L-BFGS-B", "AHR-ES", "Nelder-Mead")
par(mfrow=c(1,2), mar=c(4,4,1,1),
    oma=c(1,1,1,1))
plot(observed$prey, cex=0.75,
     ylab="prey abundance (N)", xlab="time", las=1,
     ylim=c(0,55))
lines(lbfgsb1.pred$prey, col=1, lwd=4)
lines(ahr.pred$prey, col=2, lwd=2)
lines(nm.pred$prey, col=3, lwd=2)

plot(observed$predator, cex=0.75, 
     ylab="predator abundance (P)", xlab="time", las=1,
     ylim=c(0,7))
lines(lbfgsb1.pred$predator, col=1, lwd=4)
lines(ahr.pred$predator, col=2, lwd=2)
lines(nm.pred$predator, col=3, lwd=2)

legend(100, 1.8, legend=methods, bty="n", cex=0.75, y.intersp=0.8, 
       inset=-0.0, xjust=1, pch = c(1, rep(NA,5)), lty=c(0, rep(1,5)), 
       col=c(1,1:3), lwd=2)
```

In this example, the 'L-BFGS-B' and 'AHR-ES' algorithms were able to estimate the original parameter values, but the 'Nelder-Mead' algorithm were not. 

## Fitting an EDO system: the SIR epidemiological model

As a second example, we will estimate the parameters for a SIR epidemiological model. The model is defined by a system of ordinary differential equations for the number of susceptible $S$, infected $I$ and recovered $R$ individuals:
$$\frac{dS}{dt} = -\beta S I/N$$
$$\frac{dI}{dt} = \beta S I/N -\gamma I$$
$$\frac{dR}{dt} = \gamma I$$

The parameters to estimate are the  average number of contacts per person per time $\beta$ and the instant probability of an infectious individual recovering $\gamma$. To start, we created the demonstration data for this model using the function `calibrar_demo()` function with `T=100` as an additional argument to specify the time horizon. 

```{r}
path = NULL # NULL to use the current directory
SIR = calibrar_demo(path=path, model='SIR', T=100) 
setup = calibration_setup(file = SIR$setup)
observed = calibration_data(setup=setup, path=SIR$path)
run_model = calibrar:::.SIRModel
```

To simulate the model, we will create a function taking a vector or list of parameters `par`, that solves the EDO system and return a list with the simulated variables ($S$, $I$, $R$):

```{r}
run_model = function(par, T) {
  if(!requireNamespace("deSolve", quietly = TRUE))
    stop("You need to install the 'deSolve' package.")
  # par is a list with 'alpha', 'beta' 'gamma', 'sd' and 'mu_ini'.
  SIR = function(t, y, parms, ...) {
    N = sum(unlist(parms$initial))
    beta = parms$beta
    gamma = parms$gamma
    S = y[1]
    I = y[2]
    dS = -beta*S*I/N
    dI = +beta*S*I/N -gamma*I
    dR = +gamma*I
    return(list(c(dS, dI, dR)))
  }
  times = seq(0, T)
  y0 = c(par$initial$S, par$initial$I, par$initial$R)
  sol = deSolve::ode(y=y0, times=times, func=SIR, parms=par, method="ode45")
  out = as.list(as.data.frame(sol[,-1]))
  names(out) = c("susceptible", "infected", "recovered")
  out$susceptible[is.na(out$susceptible)] = 0
  out$infected[is.na(out$infected)] = 0
  out$recovered[is.na(out$recovered)] = 0
  return(out)
}
```

As in the previous example, the objective function will be created based on the `run_model()` function and the information in the `setup` table created with the demo:

```{r, echo=FALSE, results='asis'}
knitr::kable(setup)
```

```{r}
obj = calibration_objFn(model=run_model, setup=setup, observed=observed, T=SIR$T, aggregate=TRUE)
```

Now, we can try several optimization algorithms to estimate the parameters of the model:

```{r, cache=TRUE}
lbfgsb3 = calibrate(par=SIR$guess, fn=obj, method='LBFGSB3', lower=SIR$lower, upper=SIR$upper, phases=SIR$phase)
lbfgsb2 = calibrate(par=SIR$guess, fn=obj, method='Rvmmin', lower=SIR$lower, upper=SIR$upper, phases=SIR$phase)
ahr = calibrate(par=SIR$guess, fn=obj, method='AHR-ES', lower=SIR$lower, upper=SIR$upper, phases=SIR$phase)
cg = calibrate(par=SIR$guess, fn=obj, method='Rcgmin', phases=SIR$phase)
nm = calibrate(par=SIR$guess, fn=obj, method='Nelder-Mead', phases=SIR$phase)
```

and compare them:

```{r}
summary(SIR, lbfgsb2, lbfgsb3, ahr, cg, nm, show_par = 1:2)
```

In this example, the algorithms 'Rvmmin', 'AHR-ES', 'Rcgmin' and 'Nelder-Mead' are able to estimate the original parameters, but the 'LBFGSB3' fails.

Please, refer to `vignette(package="calibrar")` for additional vignettes or to the [calibrar website](https://roliveros-ramos.github.io/calibrar/) for more details.



